---
title: "Assignment 2"
author: "Antonio Dehesa"
output: pdf_document
---

## Introduction

### Purpose

The purpose of this program is to explore the Iris dataset as the first programming assignment for the Data Mining course, perform exploratory operations, display features of the dataset, and create plots for the dataset.

### Dataset

The dataset to be used is the Iris dataset, for which more information can be found here: https://www.ritchieng.com/machine-learning-iris-dataset/


### Use case

This program is an example on the first steps to be performed on a new dataset. 

## Proposed Libraries

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)

library(datasets)
```

The required libraries for this program are: 

* datasets: This librarty includes the datasaet that we require, Iris. In this case, it is a mandatory library. 

## Dataset Analysis

We need to import the Iris dataset, for which we will explore the available information.

```{r}
data(iris)
str(iris)
```
We can see that there are 5 total variables in the dataset: Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, and Species. 
We can also see that the first four variables are of numeric type, while the last variable is of type Factor.


## Conclusions